\LARGE \bf Artifical Cognition for Social Human-Robot Interaction:\\ An Implementation

### 2016-06-16 rebuttal.tex

\documentclass{article}
\usepackage[utf8]{inputenc}

\begin{document}

Dear Editors, Dear Reviewers,

\vspace{2em}

We would like first to thank the reviewers for their thorough review of our
work. The manuscript underwent a major rewrite, impacting both the overall style
(which should now be of much higher quality) and the whole organisation of
the paper (following the reviewers' recommendations). We also dedicated time to
redraw the figures in an improved academic style.

The novelty of this contribution has been questioned by one of the reviewers.
We hopefully address this concern hereafter, along with the other comments of
Reviewer 3.

We have as well attempted to address individually each of the other concerns
raised by the reviewers in this document (note that minor typos/style issues
have been fixed directly in the manuscript, and are not discussed here).

@@

\begin{itemize}

    \item \emph{For longer journal articles, a citation style like 'apalike' that uses the
author names in the references would be more pleasant to read since it saves
reviewers from jumping back and forth all the time.}

As indicated by the Editor in Chief, we need to keep the current citation style.

    \item \emph{p9: Instead of storing these models separately (having to
    update each
of them), wouldn't it be easier to compute them once they are needed, and
to maintain a quasi-global world belief?} This is a design choice that trades
off scalability for an explicit domain. We have clarified and
justified this design choice in section 2.2 (\emph{Knowledge model}). Also, in
the specific case of agent-dependent models, we want to represent
false-beliefs situations (where one agent carries wrong beliefs on the world)
that can typically not be computed ``on demand'' because they are observable
only at specific times.

    \item \emph{p10 l 49: How is this discrimination done?} The discrimination
        algorithms are detailed in \cite{Ros2010b} (Best Paper Award at
        ROMAN2010). We now briefly mention them in the paper (p.7), and we have added a
        reference to this paper.

    \item \emph{p11 l 42-44: These concepts like "feeling uncomfortable" or "could be
      misinterpreted" are very complex and consist of multiple facets. Please
        explain which of these facets are supported by the system and how this
          information (and the classifier determining this information) is
          derived.
            I would expect the real difficulty to be how to determine these
            aspects in
              a nontrivial way, rather than their combination by considering
              them as "yet
          another cost function".}

          Section 3.4.2 attempts to better explain these techniques, based on
          \emph{a posteriori} filtering. The application of these rules is
          currently static: the execution controller calls the planner with
          pre-defined parameters. As noted by the reviewer, the online
          adaptation of these parameters, while technically straightforward,
          implies a level of interaction management that we have yet to
          explore.

    \item \emph{p15 While the production of such rules is one thing, managing the
    dynamics of this knowledge and e.g. removing it once it's outdated or updating
once it has changed are crucial. This should be explained in more detail.}
The dynamics of the knowledge are managed by the clients  of the knowledge base
themselves (ie, the deliberative components like the situation assessment
module, the dialogue processing module, etc.). In particular, the situation
assessment module recomputes at every step the whole set of spatial relations,
and update accordingly the knowledge base. This has been clarified in both
section 2.2 (\emph{Knowledge model}) and section 3.2 (\emph{Situation
Assessment, Human Awareness, Perspective Taking})

    \item \emph{The paragraphs in Section 3.4.3 appear to be rather
        disconnected, some effort should be made to turn this into a readable
        text and to explain design choices like the 3-tuple representation or the
        gripper states that just drop out of the blue.}

        This section (now 3.5) has been reworked, and should now be
        self-explanatory.

    \item \emph{The example scenarios have very bold names for often rather
        simple things.} The reviewer is indeed right. We have mistakenly used internal
        \emph{code names} for the experiments' names. These have been reworded.
        Besides, we have replaced the terminology \emph{Experiments} by
        \emph{Case studies} to better reflect the nature of these results.

    \item \emph{p 20: AFAIK KnowRob stores OWL as facts in Prolog, why
        should the TBOX be immutable?} The reviewer is right insofar as there is
        not limitations inherent to Prolog in that regard. We have removed the
        reference to KnowRob TBOX, that originated in a discussion with KnowRob's author who told
        us then that KnowRob could not update at run-time its TBox. We were
        however        not able to find literature confirming this fact, and we
        have removed this reference.

@@

Reviewer 2 raised several concerns regarding the overall article organisation
and lack of focus: \emph{``The overall impression is that the paper tries to
cove a too brad set of topics. As a result the takeaway message for the user is
too vague.''}

S/he recommended to modify the section 3 to either provide more informations
\emph{``to understand the problems and the technical solutions''} or to focus on
one particular cognitive skill; asked for \emph{``more substantial validation
of the specific problem(s) chosen in section 3''} in section 4, and suggested
several modifications to the discussion.

We have attempted to address these concerns in the following way:

\begin{itemize}
    \item section 3 underwent a large rewrite, in particular to provide
        technical insights, as well as a better overall picture of the role and
        importance of these skills. We have also added a subsection on the
        reasoning processes that take place in the background.
    \item the section 4 (previously ``experimental outcomes'') has been
        rewritten to make clear that we only present here \emph{case studies}
        that are meant to illustrate and clarify how the whole architecture
        works ``in the real world''. Besides, more details have been provided to the
        \emph{Clean the Table} study (as well as a redesigned figure) to help
        the user to follow the execution flow.
    \item the discussion section has been entirely reorganised and refocused, following the
        Reviewer recommendations. The key decisional issues have been explicitly
        stated, and overlaps between subsections have been addressed.
\end{itemize}

@@

We address hereafter the other specific points raised by the reviewer:

\begin{itemize}

    \item \emph{introduction "This article attempts however to organize them
            (i.e. the AI challenges
            in HRI) into a coherent challenge for artificial intelligence, and
            also to make explicit 
            some of the paths that have been investigated on our robots and that
            result in a deliberative, 
            knowledge-oriented, architecture for human-robot interaction." This
            suggests that the
            architecture plays a central role, but the contribution in terms of
            architecture is not 
        the most interesting one.}

        We have accordingly rephrased this paragraph to make clear that our
        contribution focuses on a set of deliberative components that interact
        to support human-robot interaction.

    \item \emph{pge 4 (section 2)
            "execution controller that may decide to take an incoming desire
            into account to create
            its own internal goals." The role of the execution controller
            remains unclear (discussed
        also in subsection 3.4.3).}

            Section 2.2 (Knowledge Model) has been largely rephrased, and
            hopefully clarifies the role of the execution controller.

    \item \emph{"Aligning the upper part of the ontology on OpenCyc ..." In my
            view the requirement of
            compatibility with other ontologies is not so critical. What would
            be really nice are
            ontologies (or techniques for extracting form general resources)
            that capture
        the common sense knowledge that is essential for robot operation.}

            We agree with the reviewer. While, until now, our focus has not been on
            merging external sources of common-sense knowledge with the robot's
            knowledge model, we believe that by initially relying on de-facto
            standards like OpenCYC ensures to some extent that we will be able
            to do so in the coming years.

    \item \emph{pg. 7 (Section 3.1) ``From an artificial intelligence
            perspective, Spark takes care of merging sensing modalities, it
            computes and grounds perceptions into symbolic knowledge (situation
            assessment), it is responsible temporal interpretation of situations
            and performs spatial perspective-taking.'' given on this
        characterization, Spark should be discussed in greater detail.}

        We have
        significantly developed the presentation of {\sc Spark}, in
        particular to exhaustively cover what and how symbolic relations are
        computed.

    \item \emph{pg 10 (section 3.2.1)
            "The system supports ...". This is one example of a topic that would
            be really interesting to the readers and it is just hinted. In
            particular, learning rules like "cats are animals" maybe really
            nice, but are such rules compatible with the existing knowledge?
            Adding them as triples may work only in absence of contradictions}

            The various natural language grounding processes that are made
            possible within this semantic architecture would require a
            relatively long discussion to be adequately explained. They are
            besides already presented in~\cite{Lemaignan2011a}.  We have added
            an explicit reference to this previous publication.

            To specifically answer the reviewer question on how possible
            contradictions are managed, we indeed check that, \emph{when
            proposed by humans}, new statements
            do not lead to contradictions before inserting
            them.

    \item \emph{pg 10 (section 3.2.1)
            "Note that, while the system benefits from the complementary
            modalities, they are not all
            required. The dialogue system can run with only the verbal modality,
            at the cost of a simpler interaction."
        Possibly true, but unjustified claim.}

        An example has been added to
        clarify this fall-back mechanism.

    \item \emph{pg. 10-11 section 3.3
            HATP planner is another interesting contribution that could be
            further detailed for the AIJ reader. 
            It is unclear why for planning a representation "outside the central
            knowledge base" is needed. What is then the relationship between the
            knowledge in the two distinct KBs?
            See also the sentence "Hence only a few ontology rules have been
            required to map both the  knowledge extracted from the situation assessment and the statements
            originated from the verbal  interaction to the planner domain." ( pg. 12)}

            The section on HATP (section 3.3) has been rewritten and extended to
            better reflect the capabilities of this planner. References to
            recent work on this tool have been added as well.  The paragraph
            that discuss the issue of procedural vs declarative knowledge has
            been rephrased as well.

    \item \emph{"The second level of integration of multi-modality is implicit:
            by computing symbolic properties 
            from the geometry, richer descriptions and hence discrimination
            possibilities are available: 
            for instance, if reachability is available, the robot may ask "do
            you mean the bottle that is 
            accessible to me?" to discriminate between the three bottles. That
            way, procedures relying on 
            discrimination transparently benefit from added modalities." Concept
            remains unclear.}

            The paragraph has been rephrased and the example
            updated to clarify the concepts.

    \item \emph{In the current form, Section 2.3 seems outside the scope of the
            paper. The discussion on working memory is quite interesting, but
            since it is still to be developed in detail it maybe a topic for
            future perspectives. In that case it would be nice to read which AI
            techniques could help tackling the issue.}

            The presentation of
            ``internal cognitive skills'' has been extended, refined and integrated
            to section 3. Specifically, several literature references have been
            added to the discussion on memory, both on the theoretical side and on
            existing implementations.

    \item \emph{``Subclasses (like Give, LookAt, etc.) are not asserted in the
            common-sense ontology, but are added by the execution controller (in
            link with the symbolic task planner) and the natural language
            processor based on what is actually performable and/or
            understandable by the robot at run-time'' This is very difficult to
            understand. Provide more details and, possibly, an example}

            The paragraph
            has been rephrased and illustrated to be easier to understand.

@@

\section{Comments from Reviewer 3}

The main concern of Reviewer 3 relates to the lack of originality of this
contribution.

The reviewer references two previous publications by the authors. \emph{Grounding
the Interaction: Anchoring Situated Discourse in Everyday Human-Robot
Interaction} has indeed been published in 2011 in the \emph{International Journal of
Social Robotics}. However, the overlap with this article appears to be minimal. The
discussion on natural language grounding, which is the focus of the previously
published article, is only briefly summarized in the present article (section
3.3) and as far as we can tell, only one picture, used as an example, as been
reused.

The second publication, \emph{When the robot considers the human...} is supposed
to appear in the proceedings of the \emph{2011 International Symposium on
Robotics Research}, but has yet to be actually published.  While this article
indeed presents our architecture in a way similar to our proposed contribution,
the fact that it belongs to the proceedings of a conference and that it has
actually not yet been published should make it mostly irrelevant.

### 2016-06-16 paper.tex
Implemented comments from reviewer 1

Human-robot interaction is a challenge for artificial intelligence. This field
lays at the crossroad of several domains of AI and requires to tackle them in a
holistic manner: modelling humans and human cognition; acquiring, representing,
manipulating in a tractable way abstract knowledge at the human level; reasoning
on this knowledge to make decisions; and eventually instantiating those
decisions into physical actions both legible to and in coordination with humans.
Many AI techniques are invited, from visual processing to symbolic reasoning,
from task planning to \emph{theory of mind} building, from reactive control to
action recognition and learning.

We do not claim to address here the issue as a whole. This article attempts
however to organise it into a coherent challenge for artificial intelligence,
and to explain and illustrate some of the paths that we have investigated on our
robots, that result in a set of deliberative, knowledge-oriented, software
components designed for human-robot interaction.

We focus on a specific class of interactions: collaborative task achievement
supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
illustrates this context: the two agents share a common space and exchange
information through multiple modalities (we specifically consider verbal
communication, deictic gestures and social gaze), and the robot is expected to achieve
interactive object manipulation, fetch and carry tasks and other similar
chores by taking into account, at every stage, the intentions, beliefs,
perspectives, skills of the human partner.  Namely, the robot must be able to
recognise, understand and participate to communication situations, both explicit
(\eg the human addresses verbally the robot) and implicit (\eg the human points
to an object); the robot must be able to take part to joint actions, both
pro-actively (by planning and proposing resulting plans to the human) and
reactively; the robot must be able to move and act in a safe, efficient and
legible way, taking into account social rules like proxemics.

@@

\subsection{Article overview}

The remaining of the article discusses the robotic architecture that we have built to
tackle the AI challenges of human-robot interaction, and how its relates to
other approaches. We propose to organise this discussion in four sections.

The next section introduces the architecture as a whole, as well as the
knowledge model that we have developed for our robots. Then,
Section~\ref{sec:impl} details each of the main components of the architecture,
and highlights their significance for artificial intelligence.
Section~\ref{sec:expe} presents two studies that illustrate in a practical way
what can be currently achieved with our robots.  The
Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally restate our
main contributions and discuss the key challenges that human-robot interaction
brings to artificial intelligence.

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
or extended with well-defined semantics. The concept of \emph{Object} represents
a typical case of ambiguous meaning: in everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. A human is
not usually considered as an object. {\sc OpenCyc} however precisely defines an
object as anything at least \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange informations between artificial
systems in a non-ambiguous way.

@@

If a human informs the robot that a given object is indeed a bottle, the robot
derives more informations on the object. Conversely, if the human affirms that
``a car is a bottle'', the reasoner may detect logical contradictions (like
inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies
on such logical consistency checks when processing natural language inputs,
both to ensure that the verbal input has been correctly acquired and parsed,
and also to verify that what the human says is logically consistent.  We
further discuss the strengths and weaknesses of this knowledge framework in
Section~\ref{krs-discussion}.

@@

\subsubsection{Working memory}
\label{memory}

Memory has been studied at length in the cognitive psychology and
neuro-psychology communities: the idea of \emph{short-term} and \emph{long-term}
memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative}
(explicit) and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985}
organises the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic}
memories into a hierarchy. Short-term memory is eventually refined with the
concept of \emph{working memory} by Baddeley~\cite{Baddeley2010}.  In the field
of cognitive architectures, the {\sc Soar} architecture~\cite{Lehman2006} is one
of those that tries to reproduce a human-like memory organisation. The GLAIR
cognitive architecture~\cite{Shapiro2009} also has a concept of long term/short
term and episodic/semantic memories.

It is worth emphasising that if memory is commonly associated to the process of
forgetting facts after a variable amount of \emph{time}, it actually covers
more mechanisms that are relevant to robotics, like priming
(concept pre-activation triggered by a specific
context~\cite{baxter2012modelling}) or reinforcement learning.

@@

We rely however on this mechanism in certain cases: some modules
like the natural language processor use the {\it short term} memory profile to
mark concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature during dialogue disambiguation for instance. On the other hand,
our perception layer does not make use of this mechanism: as described in the
next section, the environment model of the robot is continuously updated and the
derived symbolic knowledge is therefore transient: it lasts only as long as the
environment remains in the same state.

@@

\paragraph{Perspective Taking} \emph{Visual} perspective taking refers to the
ability for visually perceiving the environment from other's point of view.
This ability allows us to identify the objects in situations where the visual
perception of one person differs from the other one. In developmental
psychology, one typical example consists of two similar objects in a room (\eg
two balls) where both are visible for the child, but only one is visible for
the adult. Thus, when the adult asks the child to hand over ``the ball'', the
child is able to correctly identify which ball the adult is referring to (\ie
the one visible from the adult point of view), without asking~\cite{Moll2006}.

@@

{\sc Spark} computes three main agent-independent relations based on the
bounding box and centre of mass of the objects
(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
is on another object $O_2$, and is computed by evaluating the centre of mass of
$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
object $O_1$ is inside another object $O_2$ based on their bounding boxes
$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
is next to another object $O_2$. We do not use a simple distance threshold to
determine if two objects are next to each other since the relation is highly
dependent on the dimensions of the objects. For instance, the maximum distance
between large objects (\eg two houses) to consider them as being next to each
other is much larger than the maximum distance we would consider for two small
objects (\eg two bottles). Thus, the relation between the dimensions and the
distances of the objects are taken into account.  {\sc Spark} also compute
symbolic facts related to agent independent world dynamics.  The predicate
\concept{isMoving} states, for each tracked entity, whether it is currently
moving or not.

@@

\item \emph{Sees}: this relation describes what the agent can see, \ie what is
    within its field of view (FOV). In our current implementation, this
    affordance is computed by dynamically placing an OpenGL camera at the
    location of the eyes and running occlusion checks from it.  In
    figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
    illustrated with a grey cone (the wider one). While he is able to see the two
    small boxes on the table in front of him, the big box on his right is out of
    his FOV, and therefore, he is not able to see it.

@@

Monitoring human activity is needed by the execution controllers to track the
engagement of the human and the progress of his/her actions. Full human action
and activity recognition is a task that requires knowledge and reasoning both on
high level facts like goals, intentions and plans, as well as bottom-up data
from agent and object motions. {\sc Spark} implements a set of simple temporal and
geometric heuristics on human hand trajectories and possible objects placements
to recognize simple atomic actions. Those primitive actions are assessed through
monitoring situations like ``an empty hand is close to an object on a table''
(precursor for a \emph{pick}), or ``a hand holding an object is over a
container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
such primitives that are relied upon by the execution controller to track the
state transitions in the human plans.

@@

Natural language is a basic interaction modality that we use in our system both
as input (processing of the human speech) and as output (verbalization of the
robot intents). Natural language processing is facilitated by our architecture
mostly processing informations with semantics that are already close to the
human-level. This section presents the main features of our speech processor,
{\sc Dialogs}, that include semantic and multi-modal grounding, and interactive
disambiguation. Algorithmic and implementation details are provided
in~\cite{Lemaignan2011a}.

We acquire natural speech input from the human participants through a custom
Android-based interface that relies on the Google speech recognition API for
speech-to-text and relays the textual transcript to the robot. It is parsed into
a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
sentence are resolved with the help
of the ontology to ground concepts like objects (\ie when a user says ``pick the
can'', it resolves to which instance of \emph{can} the user is referring to) and
actions.  Sentences are classified into questions, desires and statements, and
processed accordingly.  Figure~\ref{dialogs|ex} gives a example of the
processing of a simple, non-ambiguous command.

@@

Note that, while the system benefits from the complementary modalities, they are
not all required. The dialogue system can run with the verbal modality alone, at
the cost of a simpler interaction: if the human says ``this'' without the robot
tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
possibly be available in the knowledge base, and the robot fallbacks on the
anaphora resolution step alone.

@@

A cost and a duration function are associated to each action.  The duration
function provides a duration interval for the action achievement and is used, in
one hand, to schedule the different streams and, in the other hand, as an
additional cost function.

@@

The previous sections provide a perspective on a complete deliberative
architecture for social robots, including an implementation supported
``real-world'' experimental deployments. This section synthesises next what we
see as the main challenges that human-robot interaction brings to artificial
intelligence.  We discuss first how embodied cognition is a very real challenge
in human-robot interaction; we rephrase then the requirements of joint actions
in terms of five questions; we discuss the importance of building and
maintaining a multi-level model of the human; and we finally reflect on the
importance of explicit knowledge management in robotic architectures that deal
with human-level semantics and state in that respect the current limits of our
logic framework.

@@

The challenge of symbol grounding is also tightly linked to this issue. It
corresponds to the identification or creation, and then, maintenance of a link
between the symbol (the syntactic form of knowledge that the computer
manipulates) and its semantics (its meaning, anchored in the world). The
relations between the symbol, the referent of the symbol, and mediating minds
is classically referred as the \emph{semantic triangle} and has been
extensively studied in linguistics. The issue of grounding is well known in
cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
question: ``how the semantic interpretation of a formal symbol system can be
made intrinsic to the system?''. This issue has a major practical importance in
robotic: for a robot to be both endowed with a symbolic representational and
reasoning system, and able to \emph{act} in the physical world, it must ground
its knowledge.

@@

\emph{When to perform a given action} is enriched here by the human-context
since it has to take into account the human, his/her needs, his/her rhythms and
pace, and his/her mental state. While performing its share of the task, the
robot has to produce signals directed towards the human and to respond to
signals produced back at the proper pace.

@@

The correlate of these five \emph{W-questions} is the issue the \emph{human
models}: taking appropriate decisions with and in presence of humans require
appropriate models of the human. Models that tell the robot what the human
\emph{can do}, \emph{would like to do}, \emph{knows}, \emph{could infer}, etc.

While the task of describing all the (dynamic!) human models that are useful to
robots is immense (if doable at all), we claim that it is possible to devise and
use such models in limited, but still interesting and useful, contexts such as
collaborative human-robot objects manipulation, fetch-and-carry and associated
activities in home or work environments.

@@

Motion and action execution also requires human models, and the one we use
embeds human preferences and physical constraints that need to be accounted for
when synthesizing robot motion or elaborating robot plans. This includes
proxemics (human-robot distance) and associated issues (visibility) but also
legibility and acceptability criterias expressed in terms of \emph{social rules}
that the elaborated plans should satisfy.

@@

Human-Robot Interaction is and is going to remain a challenging field for
artificial intelligence. We hope that this contribution helps with clarifying
some of these challenges.

>>>>

Human-robot interaction is a challenge for Artificial Intelligence. This field
lays at the crossroad of several domains of AI and requires to tackle them in a
holistic manner: modelling humans and human cognition; acquiring, representing,
manipulating in a tractable way abstract knowledge at the human level; reasoning
on this knowledge to make decisions; and eventually instantiating those
decisions into physical actions both legible to and in coordination with humans.
Many AI techniques are invited, from visual processing to symbolic reasoning,
from task planning to \emph{theory of mind} building, from reactive control to
action recognition and learning.

We do not claim to address here the issue as a whole. This article attempts
however to organise it into a coherent challenge for Artificial Intelligence,
and to explain and illustrate some of the paths that we have investigated on our
robots, that result in a set of deliberative, knowledge-oriented, software
components designed for human-robot interaction.

We focus on a specific class of interactions: collaborative task achievement
supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
illustrates this context: the two agents share a common space and exchange
information through multiple modalities (we specifically consider verbal
communication, deictic gestures and social gaze), and the robot is expected to achieve
interactive object manipulation, fetch and carry tasks and other similar
chores by taking into account, at every stage, the intentions, beliefs,
perspectives, skills of the human partner.  Namely, the robot must be able to
recognise, understand and participate in communication situations, both explicit
(\eg the human addresses verbally the robot) and implicit (\eg the human points
to an object); the robot must be able to take part in joint actions, both
pro-actively (by planning and proposing resulting plans to the human) and
reactively; the robot must be able to move and act in a safe, efficient and
legible way, taking into account social rules like proxemics.

@@

\subsection{Article Overview}

The remaining of the article discusses the robotic architecture that we have built to
tackle the AI challenges of human-robot interaction, and how its relates to
other approaches. We organise this discussion in four sections.

The next section introduces the architecture as a whole, as well as the
knowledge model that we have developed for our robots. Then,
Section~\ref{sec:impl} details each of the main components of the architecture,
and highlights their significance for Artificial Intelligence.
Section~\ref{sec:expe} presents two studies that illustrate in a practical way
what can be currently achieved with our robots.  The
Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally restate our
main contributions and discuss the key challenges that human-robot interaction
brings to Artificial Intelligence.

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
or extended with well-defined semantics. The concept of \emph{Object} represents
a typical case of ambiguous meaning: in everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. A human is
not usually considered as an object. {\sc OpenCyc} however precisely defines an
object as anything at least \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange information between artificial
systems in a non-ambiguous way.

@@

If a human informs the robot that a given object is indeed a bottle, the robot
derives more information on the object. Conversely, if the human affirms that
``a car is a bottle'', the reasoner may detect logical contradictions (like
inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies
on such logical consistency checks when processing natural language inputs,
both to ensure that the verbal input has been correctly acquired and parsed,
and also to verify that what the human says is logically consistent.  We
further discuss the strengths and weaknesses of this knowledge framework in
Section~\ref{krs-discussion}.

@@

\subsubsection{Working Memory}
\label{memory}

Memory has been studied at length in the cognitive psychology and
neuro-psychology communities: the idea of \emph{short-term} and \emph{long-term}
memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative}
(explicit) and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985}
organises the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic}
memories into a hierarchy. Short-term memory is eventually refined with the
concept of \emph{working memory} by Baddeley~\cite{Baddeley2010}.  In the field
of cognitive architectures, the {\sc Soar} architecture~\cite{Lehman2006} is one
of those that try to reproduce a human-like memory organisation. The GLAIR
cognitive architecture~\cite{Shapiro2009} also has a concept of long term/short
term and episodic/semantic memories.

It is worth emphasising that while memory is commonly associated with the process of
forgetting facts after a variable amount of \emph{time}, it actually covers
more mechanisms that are relevant to robotics, like priming
(concept pre-activation triggered by a specific
context~\cite{baxter2012modelling}) or reinforcement learning.

@@

We rely however on this mechanism in certain cases: some modules
like the natural language processor use the {\it short term} memory profile to
mark concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory).
Likewise, recently seen or updated geometric entities are flagged as
\concept{ActiveConcept}. We use this feature during dialogue disambiguation for
instance. On the other hand, our perception layer does not make use of this
mechanism: as described in the next section, the environment model of the robot
is continuously updated and the derived symbolic knowledge is therefore
transient: it lasts only as long as the environment remains in the same state.

@@

\paragraph{Perspective Taking} \emph{Visual} perspective taking refers to the
ability for visually perceiving the environment from another person's point of view.
This ability allows us to identify the objects in situations where the visual
perception of one person differs from the other one. In developmental
psychology, one typical example consists of two similar objects in a room (\eg
two balls) where both are visible for the child, but only one is visible for
the adult. Thus, when the adult asks the child to hand over ``the ball'', the
child is able to correctly identify which ball the adult is referring to (\ie
the one visible from the adult point of view), without asking~\cite{Moll2006}.

@@

{\sc Spark} computes three main agent-independent relations based on the
bounding box and centre of mass of the objects
(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
is on another object $O_2$, and is computed by evaluating the centre of mass of
$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
object $O_1$ is inside another object $O_2$ based on their bounding boxes
$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
is next to another object $O_2$. We do not use a simple distance threshold to
determine if two objects are next to each other since the relation is highly
dependent on the dimensions of the objects. For instance, the maximum distance
between large objects (\eg two houses) to consider them as being next to each
other is much larger than the maximum distance we would consider for two small
objects (\eg two bottles). Thus, the distance threshold is scaled with the
objects' size. {\sc Spark} also compute symbolic facts related to agent
independent world dynamics.  The predicate \concept{isMoving} states, for each
tracked entity, whether it is currently moving or not.

@@

\item \emph{Sees}: this relation describes what the agent can see, \ie what is
    within its field of view (FOV). In our current implementation, this
    affordance is computed by dynamically placing an OpenGL camera at the
    location of the eyes and running occlusion checks from it.  In
    Figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
    illustrated with a grey cone (the wider one). While he is able to see the two
    small boxes on the table in front of him, the big box on his right is out of
    his FOV, and therefore, he is not able to see it.

@@

Monitoring human activity is needed by the execution controllers to track the
engagement of the human and the progress of his/her actions. Full human action
and activity recognition is a task that requires knowledge and reasoning both on
high-level facts like goals, intentions and plans, as well as bottom-up data
from agent and object motions. {\sc Spark} implements a set of simple temporal and
geometric heuristics on human hand trajectories and possible objects placements
to recognize simple atomic actions. Those primitive actions are assessed through
monitoring situations like ``an empty hand is close to an object on a table''
(precursor for a \emph{pick}), or ``a hand holding an object is over a
container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
such primitives that are relied upon by the execution controller to track the
state transitions in the human plans.

@@

Natural language is a basic interaction modality that we use in our system both
as input (processing of the human speech) and as output (verbalization of the
robot intents). Natural language processing is facilitated by our architecture
mostly processing information with semantics that are already close to the
human-level. This section presents the main features of our speech processor,
{\sc Dialogs}, that include semantic and multi-modal grounding, and interactive
disambiguation. Algorithmic and implementation details are provided
in~\cite{Lemaignan2011a}.

We acquire natural speech input from the human participants through a custom
Android-based interface that relies on the Google speech recognition API for
speech-to-text and relays the textual transcript to the robot. It is parsed into
a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
sentence are resolved with the help
of the ontology to ground concepts like objects (\ie when a user says ``pick the
can'', it resolves to which instance of \emph{can} the user is referring to) and
actions.  Sentences are classified into questions, desires and statements, and
processed accordingly.  Figure~\ref{dialogs|ex} gives an example of the
processing of a simple, non-ambiguous command.

@@

Note that, while the system benefits from the complementary modalities, they are
not all required. The dialogue system can run with the verbal modality alone, at
the cost of a simpler interaction: if the human says ``this'' without the robot
tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
possibly be available in the knowledge base, and the robot falls back to the
anaphora resolution step alone.

@@

A cost and a duration function are associated to each action.  The duration
function provides a duration interval for the action achievement and is used, in
one hand, to schedule the different streams and, on the other hand, as an
additional cost function.

@@

The previous sections provide a perspective on a complete deliberative
architecture for social robots, including an implementation supported
``real-world'' experimental deployments. This section synthesises next what we
see as the main challenges that human-robot interaction brings to Artificial
Intelligence.  We discuss first how embodied cognition is a very real challenge
in human-robot interaction; we rephrase then the requirements of joint actions
in terms of five questions; we discuss the importance of building and
maintaining a multi-level model of the human; and we finally reflect on the
importance of explicit knowledge management in robotic architectures that deal
with human-level semantics and state in that respect the current limits of our
logic framework.

@@

The challenge of symbol grounding is also tightly linked to this issue. It
corresponds to the identification or creation, and then, maintenance of a link
between the symbol (the syntactic form of knowledge that the computer
manipulates) and its semantics (its meaning, anchored in the world). The
relation between the symbol, the referent of the symbol, and mediating minds
is classically referred as the \emph{semantic triangle} and has been
extensively studied in linguistics. The issue of grounding is well known in
cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
question: ``how the semantic interpretation of a formal symbol system can be
made intrinsic to the system?''. This issue has a major practical importance in
robotic: for a robot to be both endowed with a symbolic representational and
reasoning system, and able to \emph{act} in the physical world, it must ground
its knowledge.

@@

\emph{When to perform a given action} is enriched here by the human context
since it has to take into account the human, his/her needs, his/her rhythms and
pace, and his/her mental state. While performing its share of the task, the
robot has to produce signals directed towards the human and to respond to
signals produced back at the proper pace.

@@

The correlate of these five \emph{W-questions} is the issue the \emph{human
models}: taking appropriate decisions with and in presence of humans requires
appropriate models of the human: what the human \emph{can do}, \emph{would like
to do}, \emph{knows}, \emph{could infer}, etc.

While the task of describing all the (dynamic) human models that are useful to
robots is immense (if doable at all), we claim that it is possible to devise and
use such models in limited, but still interesting and useful, contexts such as
collaborative human-robot objects manipulation, fetch-and-carry and associated
activities in home or work environments.

@@

Motion and action execution also requires human models, and the one we use
embeds human preferences and physical constraints that need to be accounted for
when synthesizing robot motion or elaborating robot plans. This includes
proxemics (human-robot distance) and associated issues (visibility) but also
legibility and acceptability criteria expressed in terms of \emph{social rules}
that the elaborated plans should satisfy.

@@

Human-Robot Interaction is and is going to remain a challenging field for
Artificial Intelligence. We hope that this contribution helps with clarifying
some of these challenges.
